A Historical Analysis of Quark Matter Conferences: \\ Trends, Participation, and Topic Evolution

### 2025-03-19 QM/paper.tex
clean up old data, everything from single data point

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/plenary_talks_by_country.pdf}
\caption{Distribution of plenary talks by country across Quark Matter conferences from 2011 to 2025. The stacked bars show the number of plenary presentations given by speakers from each of the top 10 countries. This visualization reveals the geographical concentration of high-visibility presentations and changes in international representation over time.}
\label{fig:country_plenary}
\end{figure}

@@

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/parallel_talks_by_country.pdf}
\caption{Distribution of parallel talks by country across Quark Matter conferences from 2011 to 2025. The stacked bars show the number of parallel session presentations given by speakers from each of the top 10 countries. This visualization reveals broader participation patterns compared to plenary talks and illustrates how research contributions are distributed globally.}
\label{fig:country_parallel}
\end{figure}

@@

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/plenary_talks_by_institute.pdf}
\caption{Distribution of plenary talks by institute across Quark Matter conferences from 2011 to 2025. The stacked bars show the number of plenary presentations given by speakers from each of the top 15 institutes. This visualization reveals which research institutions have the most prominent representation in high-visibility speaking slots, highlighting potential concentration of influence in the field.}
\label{fig:institute_plenary}
\end{figure}

@@

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/parallel_talks_by_institute.pdf}
\caption{Distribution of parallel talks by institute across Quark Matter conferences from 2011 to 2025. The stacked bars show the number of parallel session presentations given by speakers from each of the top 15 institutes. This visualization reveals broader institutional participation patterns compared to plenary talks and illustrates how research contributions are distributed across research centers.}
\label{fig:institute_parallel}
\end{figure}

@@

@@

\begin{itemize}
    \item Implementing transparent selection criteria that explicitly value geographical and institutional diversity alongside scientific excellence
    \item Developing mentoring programs or speaking opportunities specifically designed to highlight researchers from underrepresented institutions and countries
    \item Rotating conference locations more systematically to encourage broader participation and reduce travel barriers
    \item Collecting and monitoring demographic data on conference participation to track progress over time
\end{itemize}

@@

\begin{itemize}
    \item Implementation of more transparent and explicitly diversity-conscious selection processes for talks, particularly plenary presentations
    \item Consideration of mechanisms to provide speaking opportunities for researchers from underrepresented countries and institutions
    \item Balancing representation of established research directions with emerging topics to ensure the conference remains at the cutting edge
    \item Improving data collection on speaker demographics to better track progress on diversity goals
    \item Developing structures to highlight work from smaller or newer research groups, potentially through dedicated sessions or presentation formats
\end{itemize}

>>>>

The analysis was implemented in Python, with dedicated scripts handling different aspects of the data pipeline. The main script (fetch\_and\_analyze\_conferences.py) processes the raw conference data and generates basic statistics and visualizations. For more complex visualizations requiring specialized processing, standalone scripts were developed, including create\_country\_trends.py for longitudinal country participation analysis and create\_institute\_bubble.py for the institutional participation bubble chart. These modular scripts ensure reproducibility and allow for targeted refinement of specific visualizations.

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/plenary_talks_by_country.pdf}
\caption{Distribution of plenary talks by country across Quark Matter conferences from 2011 to 2025. The horizontal bars show the number of plenary presentations given by speakers from each of the top 15 countries, with percentage labels indicating relative contribution. Countries are color-coded by region for better geographical context. This enhanced visualization reveals both dominant contributors and emerging participants in high-visibility presentations.}
\label{fig:country_plenary}
\end{figure}

@@

Emerging contributors such as Brazil, South Africa, and Poland are now visible in the expanded visualization, showing how the field is gradually becoming more internationally diverse even as traditional centers maintain their prominent positions.

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/parallel_talks_by_country.pdf}
\caption{Distribution of parallel talks by country across Quark Matter conferences from 2011 to 2025. The horizontal bars show the number of parallel session presentations given by speakers from each of the top 15 countries, with percentage labels indicating their share of the total presentations. This expanded visualization reveals broader participation patterns compared to plenary talks and illustrates how research contributions are distributed globally, including both established and emerging contributors.}
\label{fig:country_parallel}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/country_trends_over_time.pdf}
\caption{Trends in country representation at Quark Matter conferences from 2011 to 2025. The line graph shows the percentage of total talks given by speakers from the top 8 countries (solid lines) and selected emerging countries (dashed lines) over time. This visualization reveals both persistent patterns of representation and changing dynamics as new countries increase their participation in the field.}
\label{fig:country_trends}
\end{figure}

Figure~\ref{fig:country_trends} provides a dynamic view of how country representation has evolved over time. This visualization offers valuable insights into both established patterns and emerging trends:

The top countries show relatively stable patterns of representation, with some year-to-year fluctuations often correlated with conference location. The United States and Germany consistently maintain strong representation, reflecting their established infrastructures and research programs.

Several emerging countries show noticeable upward trends, including Brazil, Poland, and Czech Republic. These trends highlight the gradual broadening of the field's geographical base, as heavy-ion physics research expands beyond its traditional centers.

The visualization reveals interesting patterns around conference years held in particular regions, where local representation typically increases. However, these location effects appear temporary rather than leading to sustained increases in representation from host countries.

This longitudinal view complements the aggregate analyses by revealing dynamics that might be obscured in cumulative statistics. The persistent gaps between established and emerging countries suggest that while progress is being made toward greater international diversity, significant disparities remain.

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/plenary_talks_by_institute.pdf}
\caption{Distribution of plenary talks by institute across Quark Matter conferences from 2011 to 2025. The horizontal bars show the number of plenary presentations given by speakers from each of the top 20 institutes, with percentage labels indicating their relative contribution. Institutes are categorized by type (National Laboratory, University, Research Center) and sorted by total contribution. This expanded visualization reveals which research institutions have the most prominent representation in high-visibility speaking slots, highlighting potential concentration of influence in the field.}
\label{fig:institute_plenary}
\end{figure}

@@

Mid-tier institutes now visible in our expanded visualization show interesting patterns of occasional representation, often correlating with specific breakthroughs or leadership in emerging research areas. This provides context for understanding how institutes beyond the dominant few gain visibility in the field.

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/parallel_talks_by_institute.pdf}
\caption{Distribution of parallel talks by institute across Quark Matter conferences from 2011 to 2025. The horizontal bars show the number of parallel session presentations given by speakers from each of the top 20 institutes, with percentage labels indicating their share of total presentations. Institutes are categorized by type and color-coded accordingly. This comprehensive visualization reveals broader institutional participation patterns compared to plenary talks and illustrates how research contributions are distributed across a wider range of research centers.}
\label{fig:institute_parallel}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/institute_bubble_chart.pdf}
\caption{Institute contributions across Quark Matter conferences from 2011 to 2025. This bubble chart displays the presence and contribution volume of the top 30 institutes across conference years, with bubble size proportional to the number of presentations. This visualization reveals both consistently active institutions and those with intermittent but significant contributions, providing insight into how institutional participation patterns have evolved over time.}
\label{fig:institute_bubble}
\end{figure}

Figure~\ref{fig:institute_bubble} provides a novel perspective on institutional participation patterns through a time-series bubble chart. This visualization allows us to track the consistency and volume of contributions from the top 30 institutes across all conference years:

The chart reveals distinct participation patterns among institutes. Some maintain a consistent presence with relatively stable contribution volumes across all conferences, suggesting sustained research programs and established positions in the field. Others show more intermittent participation, with strong showings in specific years followed by reduced presence in others.

Major national laboratories (BNL, CERN) show some of the most consistent participation patterns, reflecting their central role in the experimental infrastructure that drives the field. University participation appears more variable, potentially reflecting shifting research priorities, funding cycles, or changes in research group composition.

Interestingly, some institutes show coordinated patterns, with increased participation in the same conference years. These correlations often align with major experimental runs or significant theoretical developments, highlighting how the field responds collectively to new opportunities and challenges.

The bubble chart also highlights emerging institutes that have increased their participation in more recent conferences, providing a window into the evolving institutional landscape of heavy-ion physics research. This longitudinal perspective complements the aggregate analyses by revealing dynamics that might be obscured in cumulative statistics.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/regional_diversity.pdf}
\caption{Regional diversity of participation at Quark Matter conferences. The left panel shows the aggregate regional distribution of all contributions, while the right panel tracks how regional representation has evolved over time. This visualization provides insight into the geographical spread of the heavy-ion physics community and highlights opportunities for increasing participation from underrepresented regions.}
\label{fig:regional_diversity}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/representation_ratio.pdf}
\caption{Representation ratio between plenary and parallel talks by country. This chart compares each country's share of high-visibility plenary talks to their share of parallel sessions, with values above 1.0 indicating overrepresentation and values below 1.0 indicating underrepresentation. The visualization highlights potential structural biases in the allocation of speaking opportunities and identifies countries that may need additional support to achieve proportional representation.}
\label{fig:representation_ratio}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figures/diversity_metrics.pdf}
\caption{Evolution of diversity metrics over time. The top panel shows the number of unique countries represented at each conference, while the bottom panel displays the Herfindahl-Hirschman Index (HHI), a measure of concentration where lower values indicate more diverse representation. Together, these metrics provide quantitative measures of how geographical diversity has changed across conference cycles.}
\label{fig:diversity_metrics}
\end{figure}

@@

\begin{itemize}
    \item Implementing transparent selection criteria that explicitly value geographical and institutional diversity alongside scientific excellence
    \item Developing mentoring programs or speaking opportunities specifically designed to highlight researchers from underrepresented institutions and countries
    \item Rotating conference locations more systematically to encourage broader participation and reduce travel barriers
    \item Collecting and monitoring demographic data on conference participation to track progress over time
    \item Implementing regional quotas or targets that ensure representation from a broader range of countries while maintaining scientific excellence
    \item Creating mentoring or speaking opportunity programs specifically for researchers from emerging countries in the field
\end{itemize}

@@

\begin{itemize}
    \item Implementation of more transparent and explicitly diversity-conscious selection processes for talks, particularly plenary presentations
    \item Consideration of mechanisms to provide speaking opportunities for researchers from underrepresented countries and institutions
    \item Balancing representation of established research directions with emerging topics to ensure the conference remains at the cutting edge
    \item Improving data collection on speaker demographics to better track progress on diversity goals
    \item Developing structures to highlight work from smaller or newer research groups, potentially through dedicated sessions or presentation formats
    \item Implementing regional quotas or targets that ensure representation from a broader range of countries while maintaining scientific excellence
    \item Creating mentoring or speaking opportunity programs specifically for researchers from emerging countries in the field
\end{itemize}
